---
title: "Phenotype Correlations"
author: "Caryn Johansen"
date: "1/8/2017"
output: html_document
---

### Introduction

Purpose: The purpose for this script is to 1) explore and visualize the phenotype and expression data
from Hirsch et al. and to create and visualize correlation matrices between each phenotype and the 
expression data.

**Paper citation:**

Hirsch CN, Foerster JM, Johnson JM, Sekhon RS, Muttoni G, Vaillancourt B, Peñagaricano F, Lindquist E, Pedraza MA, Barry K, de Leon N, Kaeppler SM, Buell CR (2014) Insights into the maize pan-genome and pan-transcriptome. The Plant Cell 26(1): 121-135. http://dx.doi.org/10.1105/tpc.113.119982

Paper is found at: http://www.plantcell.org/content/early/2014/01/31/tpc.113.119982.full.pdf+html

Project directory:

```{r}
setwd("~/Dropbox/UCD/rotations/RRI/Hirsch_correlations/")
```

### Expression Data

The FPKM data can be downloaded at http://datadryad.org/resource/doi:10.5061/dryad.r73c5

Processing data:

the downloaded data has an error-causing end-of-line quirk. Each line ends with "^I$" which is adding a final tab to each line. This is causing
the file to input into R incorrectly. 

in linux, the following command solves this error:

```{r, eval=FALSE}
sed 's/[[:blank:]]*$//' maize_503genotypes_raw_expression_values_FPKM_cleaned.txt > newfile.txt
mv newfile.txt > maize_503genotypes_raw_expression_values_FPKM_cleaned.txt
cat -vet maize_503genotypes_raw_expression_values_FPKM_cleaned.txt | less
```

use the last line to look at the end of the text lines for the '^I$', which should no longer be there.

**Data here is are just genes with expression of at least 0.5 FPKM across all genes.**

```{r}
maize_exp <- read.table("../data/processed/maize_503genotypes_raw_expression_values_FPKM_cleaned.txt", sep=" ", header=TRUE)
```

FPKM stand for Fragments Per Kilobase of transcript per Million mapped reads. Hirsch et al used
cufflinks to determine gene expression levels, and the output was expressed in FPKM as oppossed to 
RPKM. Cufflinks uses FPKM to avoid getting skewed expression values from paired-end sequencing (
sometimes the other pair of the paired end is poor quality and does not map. Or, the reads might 
be double counted. Cufflinks claims that counting fragments is more reliable and does not have these
problems)

**How do they come up with NA vs a zero?**

**What does an NA mean?**

**Are the NAs associatated with particular genes?**

need to fix this - put in all genes
```{r}
#where are the NA's? Are there patterns to where they are located?
exp.na <- maize_exp[!complete.cases(maize_exp),]
#genes with NA's, and number of NA's

na_counting_function <- function(v) {
  sum(!complete.cases(v))
}

na.info <- data.frame(gene =exp.na$gene, num.Na = apply(exp.na, 1, na_counting_function))
hist(na.info$num.Na, breaks=50)
summary(na.info)
hist(na.info$num.Na)

na.info[na.info$num.Na == 503,]
```
No, except for one.

Just want genes that are expressed in all of the samples

```{r}
maize_exp <- maize_exp[,c(1,5:507)]

maize_exp <- maize_exp[rowSums(maize_exp[,2:504] <= 0.5) == 0,]
#write.table(maize_exp, file=""data/processed/maize_503genotypes_FPKM_expressed.txt", quote = F, row.names = F, col.names = T
maize_exp <- read.table("data/processed/maize_503genotypes_FPKM_expressed.txt", sep=" ", header=TRUE)
```

**Distribution of fpkm**

**Look for list of expressed genes, Figure 1c, cufflinks confidence interval**

Just use the 18,000 genes,

average phenotypes into single number per individual, average expression

**SNP data**

Don't need SNP data yet - not for WGCNA. Need SNP data for anything 

###### FPKM distribution

```{r}
library(reshape2)

str(maize_exp)

#melt the fpkm into one column
hist(maize_exp$A797NW, breaks=1000)
exp.long <- melt(maize_exp, id="gene")

ggplot(exp.long, aes(x=value, y=..density..)) +
  geom_histogram(bins=1000) +
  geom_vline(aes(xintercept = mean(exp.long$value, na.rm=T), colour="red"))

plot(density(exp.long$value, na.rm=TRUE))

```

### Phenotype Data

if missing, pairwise deletion

There are two sets of phenotypic data, the Supplemental Data Set 6, and the dataset acquired by Jeff from the authors:

```{r}
transition <- read.table("../data/processed/tpc119982SupplementalDS6_cleaned.csv", sep=",", header=T)
phenotypes <- read.table("../data/processed/WIDIV_2010_paper_phenotypic_data_cleaned.csv", sep=",", header=T, na.strings = ".")

#for pheotype data frame, year, rep, and no should be factors:
phenotypes$Year <- as.factor(as.character(phenotypes$Year))
phenotypes$Rep <- as.factor(as.character(phenotypes$Rep))
phenotypes$No <- as.factor((phenotypes$No))

str(phenotypes)
str(transition)
```

### Phenotype summaries

Looking at the phenotypes...

The transition data show the days after planting of certain events (the last juvenile leaf, growing degree days until pollen shed, and 
growing degree days until silk)


```{r}
library(ggplot2)

transition2 <- transition[order(transition$Last.Juvenile.Leaf),]
transition2$Genotype <- factor(transition2$Genotype, levels = transition2$Genotype[order(transition2$Last.Juvenile.Leaf)])

transition_long <- melt(transition2, id.vars = "Genotype", 
                        measure.vars = c("Last.Juvenile.Leaf", "Growing.Degree.Days.to.Pollen.Shed","Growing.Degree.Days.to.Silk"), value = TRUE)

#with means
means <- data.frame(mean = colMeans(transition[,2:4], na.rm = T))
means$variable <- levels(transition_long$variable)

ggplot(data = transition_long, aes(x=Genotype, y=value)) + geom_point() +
  facet_grid(variable~., scale="free") +
  geom_hline(aes(yintercept = mean), means)


transition2 <- transition[order(transition$Growing.Degree.Days.to.Pollen.Shed),]
transition2$Genotype <- factor(transition2$Genotype, levels = transition2$Genotype[order(transition2$Growing.Degree.Days.to.Pollen.Shed)])
transition_long <- melt(transition2, id.vars = "Genotype", 
                        measure.vars = c("Last.Juvenile.Leaf", "Growing.Degree.Days.to.Pollen.Shed","Growing.Degree.Days.to.Silk"), value = TRUE)

ggplot(data = transition_long, aes(x=Genotype, y=value)) + geom_point() +
  facet_grid(variable~., scale="free") +
  geom_hline(aes(yintercept = mean), means)

```

Looking at the Phenotypes data:

```{r}
phenotypes2 <- phenotypes[,c(2,3,7:20)]
phenotypes_long <- melt(phenotypes2, id.vars = c("Entry", "Rep"), measure.vars = names(phenotypes2)[3:16])

ggplot(phenotypes_long, aes(x=variable, y=value)) + geom_boxplot() + theme(axis.text.x = element_text(angle=90, hjust=1))
```


Combined the Phenotypes into one data frame

**for phenotypes, look at them individually to transform. If they're not normal, do a log**
**transformation probably. Or, fit a model of environment (different years) and see if the**
**residuals are normally distributed. But first, just average**

```{r}
library(gplots)

traitData <- read.table("data/processed/maize_g_nT.csv", sep=",",header=T, row.names = 1)

#phenotype distribution
trait.phys <- traitData[1:17,]
trait.phys$trait <- row.names(trait.phys)
trait.phys.m <- melt(trait.phys, id.vars = "trait")
trait.phys.w <- reshape(trait.phys.m, idvar = "variable", timevar="trait", direction="wide")
#trait.phys <- t(trait.phys)

ggplot(trait.phys.m, aes(x=value, y=..density..)) +
  geom_histogram() +
  facet_wrap(~ trait, scale="free")
```

Phenotype PCA

```{r}
library(ggplot2)
phys.pc <- prcomp(na.omit(t(trait.phys)), center=TRUE)
phys.pc.x <- data.frame(phys.pc$x)
ggplot(phys.pc.x, aes(x=PC1, y=PC2)) + geom_point()

#remove NA columns
col.na <-apply(apply(trait.phys, 2, complete.cases), 2, sum)
trait.phys.c <- trait.phys[,col.na == 17]
col.na2 <- apply(apply(trait.phys.c, 2, complete.cases), 2, sum)

trait.pca <- prcomp(na.omit(trait.phys.w[,2:18]), center=TRUE, scale.=TRUE)
trait.pca.x <- as.data.frame(trait.pca$x)
gty <- trait.phys.w[complete.cases(trait.phys.w),]$variable
trait.pca.x$genotype <- gty


trait.pca <- prcomp(trait.phys.c[,1:244], center=T, scale.=TRUE)
trait.pca.x <- as.data.frame(trait.pca$x)
qplot(x=PC1, y=PC2, data=trait.pca.x, colour=rownames(trait.pca.x))
qplot(x=PC1, y=PC3, data=trait.pca.x, colour=rownames(trait.pca.x))
qplot(x=PC2, y=PC3, data=trait.pca.x, colour=rownames(trait.pca.x))
```

### Correlations between phenotypes and Gene Expression data

A correlation matrix can be used to show if there there is correlation between gene expression of certain genes
and phenotypes...

Things to think about:

To calculate a pearson correlation coefficient:

$$r = \frac{\sum(x-m_x)(y-m_y)}{\sqrt\sum(x - m_x)^2(y-m_y)^2}$$

Where x and y are two vectors of the same length, and $m_x$ and $m_y$ are the means of x and y.
(Spearman is this, but the difference between the rank of x and y variables.)

So, I think, to get a proper correlation, I would need to calculate the mean genetic value for the expression data?... Which I do not currently have the data for.

All by all correlation?

**meeting notes**

Average by line, get the single number 

48...+13 x 48...+13 matrix, looking at the peripheries

Do it pairwise....

**obs = pairwise complete**

how do phenotypes relate to the expression data.

imputing missing data?

**visualiize: historgram of r^2 values for each of the phenotypes**

**correlations among the phenotypes**

**pca on the phenotypes**

**eigen phenotypes**

**correlation with and without zeros**

**possibly transform with quantile organization, might be good for expression data get**



structure of hte gene expression and the phenotypes

How does it affect GWAS


Correlation code:

(This was done on farm)
```{r, eval=FALSE}
traitData <- read.table("data/processed/maize_g_nT.csv", sep=",",header=T, row.names = 1)

#cor on data to l
trait.cor.full <- cor(t(traitData), method="pearson", use="complete.obs")

#write this as a table
write.table(trait.cor.full, file="maize_g_nT_cor.csv", quote=FALSE, sep=",")

trait.cor.17 <- read.table("maize_g_nT_cor.csv", sep=",", header=T, row.names = 1, nrows = 17)
write.table(trait.cor.17, file="data/maize_g_nT_cor17.csv", sep=",", quote=FALSE)
```

```{r}
phy.cor <- read.table("data/processed/maize_g_nT_cor17.csv", sep=",")
#heatmap.2(as.matrix(phy.cor, trace="none", Colv = "none", Rowv=F, dendrogram = "none")

#just the physiological traits
phys <- phy.cor[,1:17]
phys2 <- phys
phys2$trait <- rownames(phys2)
phy.l <- melt(phys2, id.vars="trait")
heatmap.2(as.matrix(phys), trace="none")

ggplot(phy.l, aes(x=value, y=..density..)) +
  geom_histogram() +
  facet_wrap(~ trait, scale="fixed")

# just the genetic traits
genes.cor <- phy.cor[,18:15223]
heatmap.2(as.matrix(genes.cor), trace="none")

genes.cor2 <- genes.cor
genes.cor2$trait <- rownames(genes.cor2)
genes.l <- melt(genes.cor2, id.vars="trait")

ggplot(genes.l, aes(x=value, y=..density..)) +
  geom_histogram() +
  facet_wrap(~ trait, scale="fixed")

#ggplot(genes.l, aes(x=value, y=..density..)) +
#  geom_histogram() +
#  facet_wrap(~ variable, scale="fixed")

```

SLURM in the lab docs

Next week: slides for the data
Look through papers for next week - decide on which paper to present - more in Berg and Coop stuff?